---
title: "s6_220301_manuscript_figures"
author: "nlc"
date: "3/1/2022"
output:
  rmdformats::downcute:
      self_contained: true
      highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52, fig.retina = 3)
```

<style>
    body .main-container {
        max-width: 2160px;
    }
</style>


# Set up

## Load libraries
```{r}
library(tidyverse)
library(Seurat)
library(clusterProfiler)
library(org.Mm.eg.db)
```
## set seed

```{r}
set.seed(6)
```


## load s3_celltypes.rds

```{r}
so_merge <- readRDS('analysis_files/s4_so_merge.rds')

```

## Create cluster / celltype dictionary

```{r}
# Colors for clusters
color_dict <- so_merge@meta.data %>%
  dplyr::select(lineage, cluster_l2, celltype_full) %>%
  distinct() %>%
  arrange(lineage, cluster_l2)

color_dict$color <- scales::hue_pal()(nrow(color_dict))

color_dict <- color_dict %>%
  arrange(cluster_l2)

# Colors for lineage
lineage_color_dict <- so_merge@meta.data %>%
  dplyr::select(lineage) %>%
  distinct() %>%
  arrange(lineage)

lineage_color_dict$color <- scales::hue_pal()(nrow(lineage_color_dict))

lineage_cols <- lineage_color_dict$color
names(lineage_cols) <- lineage_color_dict$lineage

# Colors for species
species_cols <- RColorBrewer::brewer.pal(n = 3, name = 'Set1')[1:2]
names(species_cols) <- c('mouse', 'human')

```

# Figure 6

## Figure 6A - lineage & cluster umap

1200 x 600 pixels

```{r, fig.asp = 1}

p1 <- DimPlot(so_merge,
        group.by = 'lineage',
        label = FALSE,
        repel = TRUE,
        cols = lineage_color_dict$color)+
    coord_equal()+
  ggtitle('Lineage')+
  theme(plot.title = element_text(size = 25, face = 'bold'),
        legend.text = element_text(face = 'bold'))+
  guides(color = guide_legend(override.aes = list(size = 7)))

p2 <- DimPlot(so_merge,
        group.by = 'seurat_clusters',
        label = TRUE,
        label.size = 6,
        repel = FALSE,
        cols = color_dict$color)+
  coord_equal()+
  ggtitle('Unsupervised Clusters')+
  theme(plot.title = element_text(size = 25, face = 'bold'),
        legend.position = 'none')

p2+p1

```

## Figure 6B - Tumor-celltype distribution

```{r}
# Pull metadata and construct tumor<->cellfrac_type dictionary
so_merge_meta <- so_merge@meta.data

tumor_cellfrac <- so_merge_meta %>%
  dplyr::select(tumor, cellfrac_type) %>%
  distinct()

# compute lineage proportions & join with tumor<->cellfrac_type dictionary
props <- so_merge_meta %>%
  dplyr::select(lineage, tumor) %>%
  table() %>%
  prop.table(., margin = 2)

prop_gather <- as_tibble(props)

prop_gather <- full_join(x = prop_gather,
                         y = tumor_cellfrac,
                         by = 'tumor',
                         suffix = c('', ''))

f6b <- ggplot(prop_gather, aes(x = tumor, y = n, fill = lineage))+
  geom_col()+
  theme_bw()+
  facet_grid(~cellfrac_type, scales = 'free_x', space = 'free_x')+
  ylab('Fraction of tumor')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("fig6b.png", f6b, height = 3, width = 4.5, dpi = 500)
```

## Figure 6C - Distribution of unsupervised clusters (cell types)

1080 x 540

```{r}
so_meta <- so_merge@meta.data

so_meta_tumors <- so_meta %>%
  dplyr::select(celltype_full, cellfrac_type, lineage, tumor) %>%
  group_by(tumor, cellfrac_type, celltype_full, lineage) %>%
  summarize(n_cells = n()) %>%
  group_by(tumor) %>%
  mutate(freq = n_cells/sum(n_cells))

so_meta_tumors_grouped <- so_meta_tumors %>%
  group_by(cellfrac_type, celltype_full) %>%
  mutate(mean_freq = mean(freq)) %>%
  mutate(freq_error_ymin = mean(freq) - sd(freq)) %>%
  mutate(freq_error_ymax = mean(freq) + sd(freq))

so_meta_tumors_summary <- so_meta_tumors %>%
  group_by(cellfrac_type, celltype_full) %>%
  summarize(tumor_count = n(),
            mean_freq = mean(freq),
            freq_sd_ymin = mean(freq) - sd(freq),
            freq_sd_ymax = mean(freq) + sd(freq),
            freq_sem_ymin = mean(freq) - sd(freq)/(tumor_count)^0.5,
            freq_sem_ymax = mean(freq) + sd(freq)/(tumor_count)^0.5,
            lineage = lineage)

# mean freq of tumor
ggplot(so_meta_tumors_summary, aes(y = mean_freq, x = celltype_full, fill = cellfrac_type))+
  geom_col(position = 'dodge')+
  geom_point(data = so_meta_tumors, aes(x = celltype_full, y = freq), color = 'gray50', size = 1.5)+
  theme_bw()+
  geom_errorbar(aes(x = celltype_full, ymin = freq_sem_ymin, ymax = freq_sem_ymax), position = 'dodge', width = 0.5)+
  facet_grid(rows = vars(cellfrac_type), cols = vars(lineage), scales = 'free_x', space = 'free_x')+
  theme(strip.text.x = element_text(angle = 90))+
  ylab('Mean frequency \n (fraction of tumor)')+
  xlab('Celltype \n (cluster-celltype)')+
  RotatedAxis()

```

## Fig 6D: UMAP by lineage

```{r}
## epi
so_epi <- subset(so_merge, subset = lineage == 'epithelial')
so_epi <- RunUMAP(so_epi, reduction = 'iNMF', dims = 1:50)
DimPlot(so_epi,
        group.by = 'celltype_full',
        label = TRUE,
        label.size = 5)+
  coord_equal()+
  ggtitle('Epithelial')

VlnPlot(so_epi,
        features = c('Myc', 'Pten'))

## fibro
so_fibro <- subset(so_merge, subset = lineage == 'fibroblast')
so_fibro <- RunUMAP(so_fibro, reduction = 'iNMF', dims = 1:50)
DimPlot(so_fibro,
        group.by = 'celltype_full',
        label = TRUE,
        label.size = 5)+
  coord_equal()+
  ggtitle('Fibroblast')

## lym
so_lym <- subset(so_merge, subset = lineage == 'lymphoid')
so_lym <- RunUMAP(so_lym, reduction = 'iNMF', dims = 1:50)
DimPlot(so_lym,
        group.by = 'celltype_full',
        label = TRUE,
        label.size = 5)+
  coord_equal()+
  ggtitle('Lymphoid')

## mye
so_mye <- subset(so_merge, subset = lineage == 'myeloid')
so_mye <- RunUMAP(so_mye, reduction = 'iNMF', dims = 1:50)
DimPlot(so_mye,
        group.by = 'celltype_full',
        label = TRUE,
        label.size = 5,
        repel = TRUE)+
  coord_equal()+
  ggtitle('Myeloid')
```

## Figure 6E+6F: Epithelial heatmap + ego

720 x 720 heatmap

540 x 540 emapplot
```{r}
library(enrichplot)


so_epi <- subset(so_merge, subset = lineage == 'epithelial')

epi_markers <- read_csv('analysis_files/s4_epithelial_markers.csv')

top_epi_markers <- epi_markers %>%
  group_by(cluster) %>%
  filter(p_val_adj <= 0.01) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 5)

DoHeatmap(so_epi,
          features = top_epi_markers$gene)+
  ggtitle('Top epithelial genes')

DotPlot(so_epi,
        features = top_epi_markers$gene,
        group.by = 'celltype_full',
        cols = 'RdYlBu')+
  RotatedAxis()

epi_plots <- list()

for(i in unique(epi_markers$cluster)){

  # Identify current DEG
 curr_markers <- epi_markers %>%
    filter(cluster == i) %>%
  mutate(significantly_upregulated = avg_log2FC > 0.5 & p_val_adj <= 0.01)
 
 # extract significant DEG
 top_curr <- curr_markers %>%
  filter(avg_log2FC >= .5 & p_val_adj <= 0.01)
 

  ego_curr <- enrichGO(top_curr$gene,
                  OrgDb = org.Mm.eg.db,
                  keyType = 'SYMBOL',
                  ont = 'ALL')
  
  ego_curr <- pairwise_termsim(ego_curr)
  
  
  p1 <- ggplot(curr_markers, aes(x = avg_log2FC, y = -log10(p_val_adj), label = gene, color = significantly_upregulated))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()+
    ylab('Significance \n [ -log10(p_val_adj) ]')+
    xlab('Average log2 Fold Change')+
    ggtitle(paste0(i, ' vs all other epithelial'))+
    theme(legend.position = 'bottom')+
    labs(color = 'Significantly Upregulated')+
    scale_color_manual(values = c('black', 'red'))
  
  p2 <- treeplot(ego_curr)
  p3 <- dotplot(ego_curr)
  p4 <- emapplot(ego_curr,
         cex_label_category = 0.5,
         repel = TRUE,
         group_category = FALSE)+
    ggtitle(paste0('Cluster ', i, ' top ontologies'))
  
  # display plots in loop
  print(p1)
  print(p2)
  print(p3)
  print(p4)
  
  # stash plots for access later
  epi_plots[[as.character(i)]]$volcano <- p1
  epi_plots[[as.character(i)]]$tree <- p2
  epi_plots[[as.character(i)]]$dot <- p3
  epi_plots[[as.character(i)]]$emap <- p4
  
}
```


## Figure 6G+6H: Heatmap of Fibroblast markers + Enrichment plot

dotplot: 600 x 480

```{r}

so_fibro <- subset(so_merge, subset = lineage == 'fibroblast')

fibro_markers <- read_csv('analysis_files/s4_fibro_markers.csv')

top_fibro_markers <- fibro_markers %>%
  group_by(cluster) %>%
  filter(p_val_adj <= 0.01) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 10)

DoHeatmap(so_fibro,
          features = top_fibro_markers$gene)+
  ggtitle('Top fibroblast genes')

fibro_plots <- list()

for(i in unique(fibro_markers$cluster)){

  # Identify current DEG
 curr_markers <- fibro_markers %>%
    filter(cluster == i) %>%
  mutate(significantly_upregulated = avg_log2FC > 0.5 & p_val_adj <= 0.01)
 
 # extract significant DEG
 top_curr <- curr_markers %>%
  filter(avg_log2FC >= .5 & p_val_adj <= 0.01)
  
 # Enrichment of gene ontology using MSIGDB
  ego_curr <- enrichGO(top_curr$gene,
                  OrgDb = org.Mm.eg.db,
                  keyType = 'SYMBOL',
                  ont = 'ALL')
  
  ego_curr <- pairwise_termsim(ego_curr)
  
  # Generate plots
  p1 <- ggplot(curr_markers, aes(x = avg_log2FC, y = -log10(p_val_adj), label = gene, color = significantly_upregulated))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()+
    ylab('Significance \n [ -log10(p_val_adj) ]')+
    xlab('Average log2 Fold Change')+
    ggtitle(paste0(i, ' vs all other fibroblast'))+
    theme(legend.position = 'bottom')+
    labs(color = 'Significantly Upregulated')+
    scale_color_manual(values = c('black', 'red'))
  
  p2 <- treeplot(ego_curr)
  p3 <- dotplot(ego_curr)
  p4 <- emapplot(ego_curr,
         cex_label_category = 0.5,
         repel = TRUE,
         group_category = FALSE)+
    ggtitle(paste0('Cluster ', i, ' top ontologies'))
  
  # display plots in loop
  print(p1)
  print(p2)
  print(p3)
  print(p4)
  
  # stash plots for access later
  fibro_plots[[as.character(i)]]$volcano <- p1
  fibro_plots[[as.character(i)]]$tree <- p2
  fibro_plots[[as.character(i)]]$dot <- p3
  fibro_plots[[as.character(i)]]$emap <- p4
  
}

```

# Figure 7:

## Fig 7B: ScPred classification of murine clusters

900 x 800

```{r}
so_merge_scpred <- readRDS('analysis_files/s5_so_merge_subset_mda_scpred.rds')

freq <- so_merge_scpred@meta.data %>%
  dplyr::select(celltype_full, scpred_prediction) %>%
  table()

row.names(freq) <- paste0('MM: ', row.names(freq))
colnames(freq) <- paste0('HS: ', colnames(freq))

row_anno <- so_merge_scpred@meta.data %>%
  dplyr::select(lineage, celltype_full) %>%
  distinct() %>%
  mutate(celltype_full = paste0('MM: ', celltype_full))

row_anno <- data.frame(lineage = row_anno$lineage,
                       row.names = row_anno$celltype_full)

pheatmap::pheatmap(prop.table(freq, margin = 1),
                   annotation_row = row_anno,
                   main = 'scPred assignment \n Row fraction',
                   color = colorRampPalette(c('white', 'black'))(50),
                   annotation_colors = list(lineage = lineage_cols))


# ARI of matrix without epithelial cells
scpred_stroma <- so_merge_scpred@meta.data %>%
  filter(lineage != 'epithelial')

aricode::ARI(c1 = scpred_stroma$celltype_full,
             c2 = scpred_stroma$scpred_prediction)

# ARI of only epithelial cells
scpred_epi <- so_merge_scpred@meta.data %>%
  filter(lineage == 'epithelial')

aricode::ARI(c1 = scpred_epi$celltype_full,
             c2 = scpred_epi$scpred_prediction)

# Remove so to reclaim memory
rm(so_merge_scpred)
```

## Figure 7c: Integrated data iNMF score vs celltype/cluster

### Load integrated data and prepare metadata
```{r}
so_rliger <- readRDS('analysis_files/s5_xspecies_uinmf_so.rds')

so_rliger@meta.data$species <- str_split(rownames(so_rliger@meta.data), pattern = '_', simplify = TRUE)[,1]

celltype_major_l1_dict <- rbind(c('Endothelial', 'endothelial'),
                                c('CAFs', 'fibroblast'),
                                c('PVL', 'perivascular'),
                                c('B-cells', 'lymphoid'),
                                c('T-cells', 'lymphoid'),
                                c('Myeloid', 'myeloid'),
                                c('Normal Epithelial', 'epithelial'),
                                c('Plasmablasts', 'lymphoid'),
                                c('Cancer Epithelial', 'epithelial'))

# Convert celltype_major to celltype_l1

so_rliger@meta.data$celltype_major_to_l1[so_rliger@meta.data$species == 'human'] <- plyr::mapvalues(so_rliger@meta.data$celltype_major[so_rliger@meta.data$species == 'human'],
                                                     from = celltype_major_l1_dict[,1],
                                                     to = celltype_major_l1_dict[,2])

# Combine celltype_l1 from each species
so_rliger@meta.data <- tidyr::unite(data = so_rliger@meta.data,
                                    col = 'celltype_l1.5',
                                    celltype_major_to_l1,
                                    lineage,
                                    na.rm = TRUE,
                                    sep = '',
                                    remove = FALSE)

# Combine celltype_l2 and celltype_minor
so_rliger@meta.data <- tidyr::unite(data = so_rliger@meta.data,
                                    col = 'celltype_l2.5',
                                    celltype_full,
                                    celltype_minor,
                                    na.rm = TRUE,
                                    sep = '',
                                    remove = FALSE)

# Combine human 'subtype' and mouse 'cellfrac_type' 
so_rliger@meta.data <- tidyr::unite(data = so_rliger@meta.data,
                                    col = 'tumor_subtype',
                                    subtype,
                                    cellfrac_type,
                                    na.rm = TRUE,
                                    sep = '',
                                    remove = FALSE)

# Combine human 'Patient' and mouse 'cellfrac_type' 
so_rliger@meta.data <- tidyr::unite(data = so_rliger@meta.data,
                                    col = 'sample_id',
                                    Patient,
                                    tumor,
                                    na.rm = TRUE,
                                    sep = '',
                                    remove = FALSE)

# Add species label to celltype_l2.5

so_rliger@meta.data$species_celltype <- paste0(recode(so_rliger@meta.data$species,
                                                      'mouse' = 'MM: ',
                                                      'human' = 'HS: '),
                                               so_rliger@meta.data$celltype_l2.5)


species_celltype_dict <- so_rliger@meta.data %>%
  dplyr::select(species, celltype_l2.5) %>%
  distinct()



```

1200 x 1100

```{r}
# Average iNMF embedding to show relationship to cluster, annotated by celltype_l1

cluster_embedding <- tibble(seurat_clusters = so_rliger@meta.data$species_celltype) %>%
  cbind(so_rliger@reductions$inmf@cell.embeddings) %>%
  group_by(seurat_clusters) %>%
  summarize(across(everything(), list(mean)))

row_anno_tibble <- so_rliger@meta.data %>%
  dplyr::select(species_celltype, celltype_l1.5, species) %>%
  distinct()

row_anno <- data.frame(species = row_anno_tibble$species,
                       lineage = row_anno_tibble$celltype_l1.5,
                       row.names = row_anno_tibble$species_celltype)


pheatmap::pheatmap(t(scale(t(data.frame(cluster_embedding[,2:ncol(cluster_embedding)],
                              row.names = cluster_embedding$seurat_clusters)))),
                   scale = 'none',
                   main = 'Average UINMF embedding per cluster \n row z-scored',
                   annotation_row = row_anno,
                   annotation_colors = list(lineage = lineage_cols,
                                            species = species_cols))
```

## Figure 7D: Original cluster/celltype vs integrated cluster

1200 x 1200

```{r, fig.asp = 1}
clust_lineage <- so_rliger@meta.data %>%
  dplyr::select(RNA_snn_res.0.6, celltype_l1.5) %>%
  table()

clust_lineage_prop <- prop.table(clust_lineage, margin = 1)

for(i in 1:nrow(clust_lineage_prop)){
  
  if(max(clust_lineage_prop[i,]) > 0.8){
    curr_lineage <- names(which.max(clust_lineage_prop[i,]))
  }else{
    curr_lineage <- 'mixed'
  }
  
  if(i == 1){
    cluster_lineage_dict <- tibble(cluster = row.names(clust_lineage_prop)[i],
                                   consensus_lineage = curr_lineage)
  }else{
    cluster_lineage_dict <- rbind(cluster_lineage_dict,
                                  tibble(cluster = row.names(clust_lineage_prop)[i],
                                         consensus_lineage = curr_lineage))
  }
}

consensus_lineage_cols <- RColorBrewer::brewer.pal(n = length(unique(cluster_lineage_dict$consensus_lineage)), 'Set1')
names(consensus_lineage_cols) <- unique(cluster_lineage_dict$consensus_lineage)

so_rliger_meta <- so_rliger@meta.data

freq_table <- so_rliger_meta %>%
  dplyr::select(species_celltype, RNA_snn_res.0.6) %>%
  table()


row_anno <- so_rliger_meta %>% 
  dplyr::select(species_celltype, species, celltype_l1.5) %>%
  distinct()

row_anno <- data.frame(species = row_anno$species,
                       lineage = row_anno$celltype_l1.5,
                       row.names = row_anno$species_celltype)

row_anno$lineage <- factor(row_anno$lineage, levels = unique(cluster_lineage_dict$consensus_lineage))

col_anno <- data.frame(row.names = cluster_lineage_dict$cluster,
                       consensus_lineage = cluster_lineage_dict$consensus_lineage)

anno_colors <- list(lineage = lineage_cols,
                    consensus_lineage = consensus_lineage_cols,
                    species = species_cols)

col_order <- cluster_lineage_dict %>%
  arrange(consensus_lineage, cluster)

pheatmap::pheatmap(prop.table(freq_table, margin = 1),
                   annotation_row = row_anno,
                   annotation_col = col_anno,
                   annotation_colors = anno_colors,
                   color = colorRampPalette(colors = c('white', 'black'))(100),
                   main = 'Original identity versus integrated cluster \n (Row fraction)',
                   cluster_cols = TRUE)

```

## Figure 7E: Integrated UMAP colored by lineage

```{r}
DimPlot(so_rliger,
        group.by = 'celltype_l1.5',
        label = FALSE,
        repel = TRUE,
        cols = lineage_color_dict$color)+
    coord_equal()+
  ggtitle('Lineage')+
  theme(plot.title = element_text(size = 25, face = 'bold'),
        legend.text = element_text(face = 'bold'))+
  guides(color = guide_legend(override.aes = list(size = 7)))
```


## Figure 7F: UMAP split by species
480 x 480 per umap

```{r}
# Randomly subset to an equivalent number

species_count <- table(so_rliger@meta.data$species)

mouse_barcodes <- so_rliger@meta.data %>%
  mutate(cell_id = rownames(so_rliger@meta.data)) %>%
  filter(species == 'mouse') %>%
  dplyr::select(cell_id) %>%
  pull()

set.seed(6)
human_barcodes <- so_rliger@meta.data %>%
  mutate(cell_id = rownames(so_rliger@meta.data)) %>%
  filter(species == 'human') %>%
  mutate(random_n = sample(species_count[1])) %>%
  filter(random_n <= species_count[2]) %>%
  dplyr::select(cell_id) %>%
  pull()

so_rliger_subset <- subset(so_rliger, cells = c(mouse_barcodes, human_barcodes))

DimPlot(so_rliger_subset,
        group.by = 'RNA_snn_res.0.6',
        label = TRUE,
        label.size = 5,
        split.by = 'species')+
    coord_equal()+
    theme(legend.position = 'none')+
  ggtitle('Unsupervised clusters \n (subset to equal n)')

```

# Supplementary Figure 13:

## S13D: Integrated scRNAseq Lineage vs fraction of cluster

1200 x 300

```{r, fig.asp = 1}
clust_lineage <- so_rliger@meta.data %>%
  dplyr::select(RNA_snn_res.0.6, celltype_l1.5) %>%
  table()

clust_lineage_prop <- prop.table(clust_lineage, margin = 1)

pheatmap::pheatmap(t(clust_lineage_prop),
                   display_numbers = round(t(clust_lineage_prop),2),
                   main = 'Fraction of lineage per cluster',
                   color = colorRampPalette(colors = c('white', 'black'))(100))

for(i in 1:nrow(clust_lineage_prop)){
  
  if(max(clust_lineage_prop[i,]) > 0.8){
    curr_lineage <- names(which.max(clust_lineage_prop[i,]))
  }else{
    curr_lineage <- 'mixed'
  }
  
  if(i == 1){
    cluster_lineage_dict <- tibble(cluster = row.names(clust_lineage_prop)[i],
                                   consensus_lineage = curr_lineage)
  }else{
    cluster_lineage_dict <- rbind(cluster_lineage_dict,
                                  tibble(cluster = row.names(clust_lineage_prop)[i],
                                         consensus_lineage = curr_lineage))
  }
}


```



## Patient breakdown

```{r}
meta_human <- so_rliger@meta.data %>%
  filter(species == 'human') %>%
  mutate(og_label = celltype_minor) %>%
  mutate(sample_id = Patient)

meta_human$celltype_l1 <- plyr::mapvalues(x = meta_human$celltype_major,
                                          from = celltype_major_l1_dict[,1],
                                          to = celltype_major_l1_dict[,2])

meta_mouse <- so_rliger@meta.data %>%
  filter(species == 'mouse') %>%
  mutate(og_label = celltype_full) %>%
  mutate(sample_id = tumor)

meta <- rbind(meta_mouse, meta_human)

freq_table <- meta %>%
  dplyr::select(RNA_snn_res.0.9, sample_id) %>%
  table()

col_anno <- meta %>%
  dplyr::select(sample_id, subtype, cellfrac_type, species) %>%
  distinct()

col_anno <- data.frame(subtype = col_anno$subtype,
                       cellfrac_type = col_anno$cellfrac_type,
                       species = col_anno$species,
                       row.names = col_anno$sample_id)

pheatmap::pheatmap(prop.table(freq_table, margin = 2),
                   annotation_col = col_anno)
```
# Supplementary figure 13



## S13A - UMAP comparison - no integration, harmony and iNMF

all: 500 x 500

```{r}
# Harmony
so_harmony <- RunUMAP(so_merge, dims = 1:50, reduction = 'harmony')
harmony_umap <- DimPlot(so_harmony,
                        group.by = 'tumor')

rm(so_harmony)

# No integration
so_noint <- RunPCA(so_merge)
so_noint <- RunUMAP(so_noint, dims = 1:50, reduction = 'pca')
noint_umap <- DimPlot(so_noint,
                      group.by = 'tumor')+
  coord_equal()
rm(so_noint)

# iNMF
inmf_umap <- DimPlot(so_merge,
                     group.by = 'tumor')+
  coord_equal()


noint_umap+
  ggtitle('No integration \n 50 PCs')

harmony_umap+
  ggtitle('Harmony integrated \n 50 components')

inmf_umap+
  ggtitle('rLiger integrated \n 50 Factors')

```

## S13B - clustering optimization

Created in s3_clustering_optimization.rmd

## S13C - cluster vs lineage marker dotplot

 1080x480 pixels

```{r}
DotPlot(so_merge,
        features = c('Ptprc', 'Cd74', 'Epcam', 'Cd14', 'Cd3e', 'Pdgfra', 'Krt5', 'Pecam1', 'Mki67', 'Pdgfrb'),
        cols = 'RdYlBu',
        group.by = 'cluster_l2')+
  theme_bw()+
  coord_flip()
```

## S13D/E - c14 subset: umap & cluster vs lineage marker dotplot

480 x 480 both

```{r}
library(harmony)
c14 <- subset(so_merge,
              subset = seurat_clusters == 14)

c14 <- FindVariableFeatures(c14)
c14 <- NormalizeData(c14)
c14 <- ScaleData(c14)
c14 <- RunPCA(c14)
c14 <- RunHarmony(c14,
                        group.by.vars = 'library_id')

ElbowPlot(c14, reduction = 'harmony')

c14 <- RunUMAP(c14, reduction = 'harmony', dims = 1:5,
                  seed.use = 500)

DimPlot(c14,
        label = TRUE,
        label.size = 6)+
  ggtitle('Cluster 14 \n (5 Harmony components)')

DotPlot(c14,
        features = c('Ptprc', 'Cd74', 'Epcam', 'Cd14', 'Cd3e', 'Pdgfra', 'Krt5', 'Pecam1', 'Pdgfrb'),
        cols = 'RdYlBu',
        group.by = 'cluster_l2')+
  theme_bw()+
  coord_flip()
```

## S13F: UMAP by cellfrac_type

1080x540

```{r}
# Find smallest number of cells per cellfrac type
n_per_cellfrac_type <- so_merge@meta.data %>%
  group_by(cellfrac_type) %>%
  summarize(n_cells = n())

n_smallest <- min(n_per_cellfrac_type$n_cells)

# Randomly subsample each cellfrac type

set.seed(7)

sp_barcodes <- so_merge@meta.data %>%
  mutate(cell_id = rownames(so_merge@meta.data)) %>%
  filter(cellfrac_type == 'SP') %>%
  mutate(random_n = sample(nrow(.))) %>%
  filter(random_n <= n_smallest) %>%
  dplyr::select(cell_id) %>%
  pull()

set.seed(8)

srip_barcodes <- so_merge@meta.data %>%
  mutate(cell_id = rownames(so_merge@meta.data)) %>%
  filter(cellfrac_type == 'SR_IP') %>%
  mutate(random_n = sample(nrow(.))) %>%
  filter(random_n <= n_smallest) %>%
  dplyr::select(cell_id) %>%
  pull()

set.seed(9)

srir_barcodes <- so_merge@meta.data %>%
  mutate(cell_id = rownames(so_merge@meta.data)) %>%
  filter(cellfrac_type == 'SR_IR') %>%
  mutate(random_n = sample(nrow(.))) %>%
  filter(random_n <= n_smallest) %>%
  dplyr::select(cell_id) %>%
  pull()

# Plot UMAP by cellfrac type

DimPlot(subset(so_merge, cells = c(sp_barcodes, srip_barcodes, srir_barcodes)),
        group.by = 'lineage',
        label = FALSE,
        split.by = 'cellfrac_type',
        pt.size = 1.5)+
    coord_equal()+
    theme(legend.position = 'bottom')+
  ggtitle('Lineage by cellfrac_type \n (subset to equal n)')+
  guides(color = guide_legend(nrow = 1, override.aes = list(size = 5)))

```

## Supplemental 13G

960 x 480 pixels

```{r}
so_merge_meta <- so_merge@meta.data %>%
  mutate(tumor_pheno = paste0(tumor, '-', histology))

freq_table <- so_merge_meta %>%
  dplyr::select(lineage, tumor_pheno) %>%
  table()

cellfrac_type_dict <- so_merge_meta %>%
  dplyr::select(cellfrac_type, tumor_pheno) %>%
  distinct()

col_anno <- tibble(histology = str_split(colnames(freq_table), pattern = '-', simplify = TRUE)[,2],
                       log10_n_cells = log10(colSums(freq_table)),
                       tumor_pheno = colnames(freq_table))

col_anno <- full_join(x = col_anno,
                       y = cellfrac_type_dict,
                       by = 'tumor_pheno',
                       suffix = c('', ''))

col_anno <- data.frame(histology = col_anno$histology,
                       cellfrac_type = col_anno$cellfrac_type,
                       log10_n_cells = col_anno$log10_n_cells,
                       row.names = col_anno$tumor_pheno)


cellfrac_cols <- scales::hue_pal()(3)
names(cellfrac_cols) <- c('SP', 'SR_IP', 'SR_IR')

anno_colors <- list(cellfrac_type = cellfrac_cols,
                   histology = c(SP = 'magenta', SR = 'cyan'))

pheatmap::pheatmap(prop.table(freq_table, margin = 2), 
                   display_numbers = freq_table,
                   main = 'column scaled (cell fraction) \n Number = n_cells per celltype per tumor',
                   annotation_col = col_anno,
                   annotation_colors = anno_colors)
```

# Supplementary Figure 14

## S14A - Top DEG dotplot

1920 x 720

```{r, fig.asp = 0.5}
markers <- read_csv('analysis_files/s3_cluster_l2_markers.csv')

top_markers <- markers %>%
  filter(p_val_adj <= 0.01) %>%
  filter(pct.1 >= 0.5) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 3)

DotPlot(so_merge,
        features = unique(top_markers$gene),
        cols = 'RdYlBu',
        group.by = 'celltype_full')+
  RotatedAxis()

```

## S14B/C -  NMF score vs lineage/cluster

lineage vs iNMF embedding: 1080 x 480

cluster vs iNMF embedding: 1080 x 600

```{r, fig.asp = 1}
# Average iNMF embeddings to show relationship to celltype_l1
cluster_embedding <- tibble(lineage = so_merge@meta.data$lineage) %>%
  cbind(so_merge@reductions$iNMF@cell.embeddings) %>%
  group_by(lineage) %>%
  summarize(across(everything(), list(mean)))

pheatmap::pheatmap(t(scale(t(data.frame(cluster_embedding[,2:ncol(cluster_embedding)],
                              row.names = cluster_embedding$lineage)))),
                   scale = 'none',
                   main = 'Average iNMF embedding per cluster \n Z-scored by row')

# Average iNMF embedding to show relationship to cluster, annotated by celltype_l1
cluster_embedding <- tibble(celltype_full = so_merge@meta.data$celltype_full) %>%
  cbind(so_merge@reductions$iNMF@cell.embeddings) %>%
  group_by(celltype_full) %>%
  summarize(across(everything(), list(mean)))

annotation_rows <- so_merge@meta.data %>%
  dplyr::select(lineage, celltype_full) %>%
  group_by_all() %>%
  distinct()

annotation_rows <- data.frame(celltype = annotation_rows$lineage,
                              row.names = annotation_rows$celltype_full)

pheatmap::pheatmap(t(scale(t(data.frame(cluster_embedding[,2:ncol(cluster_embedding)],
                              row.names = cluster_embedding$celltype_full)))),
                   scale = 'none',
                   main = 'Average iNMF embedding per cluster \n Z-scored by row',
                   annotation_row = annotation_rows)

```

## S14D: Fraction of tumor by lineage (only lineages with >1 cluster)

```{r}

so_meta <- so_merge@meta.data

subtype_counts <- so_meta %>%
  dplyr::select(tumor, cellfrac_type) %>%
  distinct() %>%
  group_by(cellfrac_type) %>%
  summarize(n_tumors = n())

so_meta_tumors <- so_meta %>%
  filter(lineage %in% c('epithelial', 'fibroblast', 'lymphoid', 'myeloid')) %>%
  dplyr::select(celltype_full, cellfrac_type, lineage, tumor) %>%
  group_by(tumor, cellfrac_type, celltype_full, lineage) %>%
  summarize(n_cells = n()) %>%
  group_by(tumor, lineage) %>%
  mutate(freq = n_cells/sum(n_cells))

so_meta_tumors_grouped <- full_join(x = so_meta_tumors,
                                    y = subtype_counts,
                                    by = 'cellfrac_type') %>%
  group_by(cellfrac_type, celltype_full) %>%
  mutate(mean_freq = mean(freq)) %>%
  mutate(freq_sem_ymin = mean(freq) - sd(freq)/(n_tumors^0.5)) %>%
  mutate(freq_sem_ymax = mean(freq) + sd(freq)/(n_tumors^0.5))

so_meta_tumors_summary <- so_meta_tumors %>%
  group_by(cellfrac_type, celltype_full) %>%
  summarize(tumor_count = n(),
            mean_freq = mean(freq),
            freq_sd_ymin = mean(freq) - sd(freq),
            freq_sd_ymax = mean(freq) + sd(freq),
            freq_sem_ymin = mean(freq) - sd(freq)/(tumor_count)^0.5,
            freq_sem_ymax = mean(freq) + sd(freq)/(tumor_count)^0.5,
            lineage = lineage)

# mean freq of tumor
ggplot(so_meta_tumors_summary, aes(y = mean_freq, x = celltype_full, fill = cellfrac_type))+
  geom_col(position = 'dodge')+
  geom_point(data = so_meta_tumors, aes(x = celltype_full, y = freq), color = 'gray50', size = 1.5)+
  theme_bw()+
  geom_errorbar(aes(x = celltype_full, ymin = freq_sem_ymin, ymax = freq_sem_ymax), position = 'dodge', width = 0.5)+
  facet_grid(rows = vars(cellfrac_type), cols = vars(lineage), scales = 'free_x', space = 'free_x')+
  theme(strip.text.x = element_text(angle = 90))+
  ylab('Mean frequency \n (fraction of tumor)')+
  xlab('Celltype \n (cluster-celltype)')+
  RotatedAxis()

```

## S14D - Cluster distribution per tumor

1600 x 720

```{r}

so_meta <- so_merge@meta.data

so_meta_tumors2 <- so_meta %>%
  dplyr::select(celltype_full, cellfrac_type, lineage, tumor) %>%
  group_by(tumor, cellfrac_type, celltype_full, lineage) %>%
  summarize(n_cells = n()) %>%
  group_by(tumor) %>%
  mutate(freq = n_cells/sum(n_cells))

# Order tumors by subtype
tumor_order <- so_meta_tumors2 %>%
  ungroup() %>%
  dplyr::select(tumor, cellfrac_type) %>%
  distinct() %>%
  arrange(cellfrac_type) %>%
  dplyr::select(tumor) %>%
  pull()

# Add tumor order as factor to tumor id & find frequency of cluster per tumor
so_meta_tumors2 <- so_meta_tumors2 %>%
  mutate(tumor = factor(tumor, levels = tumor_order)) %>%
  group_by(tumor) %>%
  mutate(freq = n_cells/sum(n_cells))


# Plot
ggplot(so_meta_tumors2, aes(x = freq, y = celltype_full, fill = cellfrac_type))+
  geom_col()+
  theme_bw()+
  facet_grid(lineage~tumor, scales = 'free_y', space = 'free')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.y = element_text(angle = 0))+
  theme(plot.margin = unit(c(1,1,1,1), 'cm'))+
  ylab('Cluster')+
  xlab('Frequency of cluster in tumor')

```

## S14E: Epithelial biomarker expression

```{r}
so_epi <- subset(so_merge,
                 subset = lineage == 'epithelial')

Idents(so_epi) <- 'celltype_full'

krts <- grep(rownames(so_epi),
             pattern = '^Krt',
             value = TRUE)

goi <- c(krts,
         'Epcam',
         'Vim',
         'Cd24',
         'Cd44')

goi_exp <- AverageExpression(so_epi,
                             features = goi)$RNA

goi_exp <- goi_exp[rowSums(goi_exp) > 1, ]

pheatmap::pheatmap(t(goi_exp),
                   color = colorRampPalette(c('purple', 'black', 'yellow'))(50),
                   main = 'Luminal-basal marker expression',
                   scale = 'column')
```

# Supplemental Figure 15

## S15A: UINMF jaccard index

560 x 860 

```{r, fig.asp = 1}
top_n <- 50

n_shared_factors <- ncol(so_rliger@reductions$inmf@feature.loadings)


inmf_top_features <- list()

for(i in 1:n_shared_factors){
  curr_features <- so_rliger@reductions$inmf@feature.loadings[,i]
  
  top_features <- sort(curr_features, decreasing = TRUE)[1:top_n]
  
  inmf_top_features[[i]] <- names(top_features)
  
}


inmf_jaccard <- array(dim = c(n_shared_factors, n_shared_factors))
rownames(inmf_jaccard) <- colnames(inmf_jaccard) <- 1:n_shared_factors

for(i in 1:n_shared_factors){
  for(j in 1:n_shared_factors){
    inmf_jaccard[i,j] <- length(intersect(inmf_top_features[[i]], inmf_top_features[[j]]))/length(union(inmf_top_features[[i]], inmf_top_features[[j]]))
    
  }
}

pheatmap::pheatmap(inmf_jaccard,
                   color = colorRampPalette(colors = c('white', 'red'))(100),
                   breaks = seq(from = 0, to = 0.5, by = 0.005),
                   main = 'Jaccard similarity of UINMF shared-feature factor loadings',
                   cutree_rows = 7,
                   cutree_cols = 7)

```

## S15B: Average UINMF embedding per lineage

```{r}
lineage_embedding <- tibble(lineage = so_rliger@meta.data$celltype_l1.5) %>%
  cbind(so_rliger@reductions$inmf@cell.embeddings) %>%
  group_by(lineage) %>%
  summarize(across(everything(), list(mean)))

pheatmap::pheatmap(scale(t(data.frame(lineage_embedding[,2:ncol(lineage_embedding)],
                              row.names = lineage_embedding$lineage))),
                   scale = 'none',
                   main = 'Average UINMF embedding per lineage \n row z-scored',)

```

## S15C: Clustering optimization

Figure produced in script S5_human_tnbc_integration_rliger

## S15D: Fraction of lineage per cluster

1200 x 300

```{r}
clust_lineage <- so_rliger@meta.data %>%
  dplyr::select(RNA_snn_res.0.6, celltype_l1.5) %>%
  table()

clust_lineage_prop <- prop.table(clust_lineage, margin = 1)

pheatmap::pheatmap(t(clust_lineage_prop),
                   display_numbers = round(t(clust_lineage_prop),2),
                   main = 'Fraction of lineage per cluster',
                   color = colorRampPalette(colors = c('white', 'black'))(100))
```

## S15E: Sample composition vs unsupervised cluster

```{r}
freqs <- so_rliger_meta %>%
  dplyr::select(sample_id, seurat_clusters) %>%
  table()

col_anno <- data.frame(row.names = cluster_lineage_dict$cluster,
                       lineage = cluster_lineage_dict$consensus_lineage)

row_anno <- so_rliger_meta %>%
  dplyr::select(sample_id, species, tumor_subtype) %>%
  distinct()

row_anno <- data.frame(row_anno[,-1],
                       row.names = row_anno$sample_id)

anno_colors <- list(lineage = lineage_cols)

col_order <- cluster_lineage_dict %>%
  arrange(consensus_lineage, cluster)

pheatmap::pheatmap(prop.table(freqs, margin = 1),
                   annotation_col = col_anno,
                   annotation_row = row_anno,
                   color = colorRampPalette(colors = c('white', 'black'))(100),
                   main = ('Sample composition \n Sample ID vs unsupervised cluster \n (Row Fraction)'))
```

# sessionInfo()

```{r}
sessionInfo()
```

